\textbf{AI assistants}: A framework for semi-automated, transparent, tooling-rich data wrangling

### 2018-11-02 ai-assistants/paper.tex
More work on AI assistants report

\documentclass{article}
\usepackage{color}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{tgtermes}
\usepackage{forest}
\usepackage{fancyvrb}

\forestset{
  tt/.style={
    for tree={
      parent anchor=south,
      child anchor=north,
      grow=270,
      align=center,
      inner sep=0pt
    }
  },
  nice empty nodes up/.style={
    for tree={calign=fixed edge angles},
    delay={where content={}{shape=coordinate,for parent={for children={anchor=south}}}{}}
  }
}
\begin{document}
\title{What is an AI assistant?}
\author{Tomas Petricek}
\maketitle

\newcommand{\ident}[1]{\textnormal{\sffamily #1}}

\section{Introduction}
The idea of an \emph{AI assistant} has been discussed in many contexts in the AIDA project.
The Wrattler notebook system aims to provide a platform through which AI assistants can be 
easily used; the datadiff project developed an AI assistant which automates the process of 
reconciling inconsistencies between pairs of tabular datasets. However, we never clearly defined
what an AI assistant is. This document aims to collect some notes to help answer this question.
In the current version, these are mostly based on the needs of Wrattler and datadiff integration
in Wrattler -- as such, it clearly does not give the final answer, but merely one (limited, but
hopefully useful) perspective.

A simple way of trying to address the data wrangling problem would be to build a set of functions
that take some dirty datasets and produce one or more clean datasets. The user could then call these
functions on their dirty data and would get a clean result (conceivably, this could also be further 
automated). However, this simple approach has two problems. First, real-world data sets often 
contain numerous corner cases that require some human insight. Second, we want the data analysis
to be transparent and be able to understand how have the data wrangling problems been solved. 
A simple function does not address these because it does not encourage human interaction\footnote{At
best, users can run the function again with different parameters, but they get no suggestion about 
suitable values of the parameters.} and it is not transparent -- it produces a result without 
indicating how and why.

\section{What makes AI assistants unique}

AI assistants address the above issues as follows. First, AI assistants are interactive. Given 
some input data, they perform some analysis and provide the user with several options to choose 
from (those can be most likely solutions to some problem, possible ways of resolving ambiguous
corner cases, or a choice of next steps the AI assistant can follow). The options can be a finite
list (where the user has to choose one) or open-ended (where the user can specify some input).
They should also be sorted by likelihood (if we always choose the most likely option, the AI
assistant will behave as a fully automatic function).

@@

Note that each AI assistant can use its own domain-specific language -- the language should be
as small as possible, i.e. it should be able to express the data transformations that the AI
assistant can recommend, but nothing more.

The combination of interactivity and domain-specific languages allows us to do a number of 
interesting things:

@@

\section{AI assistants as trees}
We can think of an AI assistant as a function that takes some input data and generates
a tree where each node offers a finite list of options that the user can choose from. (For
simplicity, we ignore the possibility of open-ended options where the user specifies some input 
and we also ignore the fact that the AI assistant might take some initial parameters in addition
to the input). Formally:
%
\begin{equation*}
\begin{array}{rcll}
a &=& \ident{Data} \rightarrow t &\textnormal{(AI assistant)}\\
t &=& (n_1, p_1, e_k, t_1), \ldots (n_k, p_k, e_k, t_k) &\textnormal{(Provided options)}\\
\mathit{eval} &=& \ident{Data}\times e \rightarrow \ident{Data}&\textnormal{(Evaluator for the DSL)}
\end{array}
\end{equation*}
%
An AI assistant $a$ is a function that takes some input data and returns a root tree node $t$. 
A tree node contains a list of options $(n_i, p_i, e_i, t_i)$ where $n_i$ is a name of the node (shown
to the user, but otherwise unimportant), $p_i$ is a number that gives the likelihood of the 
option (between 0 and 1), $e_i$ is an expression in the domain-specific language that represents
the data transformation attached to the node and $t_i$ is a sub-tree that can be empty or can offer 
more follow-up options. Note that this sub-tree can be generated lazily -- it is only needed if the 
user chooses the option with which it is associated and so we do not need to evaluate the full tree 
upfront.

The nature of expressions $e$ depends on the particular AI assistant and so we leave that abstract
for now. However, we certainly need to be able to apply the transformation on sample dataset, 
which is captured by the $\mathit{eval}$ function, which takes data together with a DSL expression
and returns transformed data. An (abstract) example tree might look as follows:

\vspace{0.5em}
\begin{forest}, tt  
  [
    [{$\mathbf{n_1}$\\$p=0.9$\\$e=e_1$}
      [{$\quad\mathbf{n_{11}}\quad$\\$p=0.9$\\$e=e_{11}$}]
      [{$\quad\mathbf{n_{12}}\quad$\\$p=0.7$\\$e=e_{12}$}]
    ]
    [{$\mathbf{n_2}$\\$p=0.8$\\$e=e_2$}
      [{$\quad\mathbf{n_{21}}\quad$\\$p=0.9$\\$e=e_{21}$}]
      [{$\quad\mathbf{n_{22}}\quad$\\$p=0.8$\\$e=e_{22}$}]
    ]
    [{$\mathbf{n_3}$\\$p=0.7$\\$e=e_3$}
      [{$\quad\mathbf{n_{31}}\quad$\\$p=0.8$\\$e=e_{31}$}]
      [{$\quad\mathbf{n_{32}}\quad$\\$p=0.6$\\$e=e_{32}$}]
    ]
  ]
\end{forest}
\vspace{0.5em}

\noindent
Here, the root offers three options represented by nodes $n_1, n_2, n_3$. The most likely one
as recommended by the AI assistant is $n_1$. The expression attached to the node is $e_1$ and
when the user selects the node, we can display a preview using sample data $\mathit{sample}$
by evaluating $\mathit{eval}(\mathit{sample}, e_1)$.

Assuming the user chooses the most likely option $n_1$, they will then be presented with two
subsequent options, $n_{11}, n_{12}$. Even though the AI assistant marks the first node as more
likely, the user may review the previews $\mathit{eval}(\mathit{sample}, e_{11})$,
$\mathit{eval}(\mathit{sample}, e_{12})$ and decide that the latter is actually the correct
data cleaning operation. They can then choose the node $n_{12}$ as the final one use the 
cleaning script represented by the DSL expression $e_{12}$.

\section{Examples of AI assistants}

What would a sample AI assistant look like when seen through the above model? We consider two
examples. The first is hypothetical AI assistant for extracting data from CSV files. The second
is the datadiff assistant. For datadiff, we discuss the current implementation in Wrattler, but
also briefly sketch a possible improvement (that improves the interactivity aspect).

@@

As an example, consider the following broken CSV file. The fle starts with two meta-data lines
(one specifying the name and another specifying copyright information). This is followed by headers
and actual data.

@@

A hypothetical minimalistic AI assistant for cleaning CSV files could generate tree with two
levels of nodes. In the first level, it will aim to infer where the actual data is in the CSV
file (and it will offer its best guesses to the user). In the second level, it will aim to infer
the types of the columns.

The domain-specific language of data cleaning operations consists of two functions named
$\ident{range}$ and $\ident{types}$ that specify the range of data in the file and the types
of the columns: $\ident{range}(\ident{lines}=l_n\ldots l_m, \ident{cols}=k)$ denotes that
the data starts on line $l_n$ and ends on line $l_m$ and there are $k$ columns; 
$\ident{types}(t_1,\ldots, t_k)$ specifies that the types of the $k$ columns are 
$t_1, \ldots, t_k$. The types are either $\ident{string}$ or $\ident{int}$.

Given the above file as an input, the AI assistant might produce the following tree:
%
\begin{equation*}
\begin{array}{l}
p=0.9,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3)\\[0.2em]
\qquad\longrightarrow p=0.9,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{string})\\[0.2em]
\qquad\longrightarrow p=0.3,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{int})\\[0.4em]
p=0.7,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3)\\[0.2em]
\qquad\longrightarrow p=0.8,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{int})\\[0.2em]
\qquad\longrightarrow p=0.7,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{string})\\[0.4em]
p=0.3,~ e=\ident{range}(\ident{lines}=1\ldots 6, \ident{cols}=1)\\[0.2em]
\qquad\longrightarrow p=0.9,~ e=\ident{range}(\ident{lines}=1\ldots 6, \ident{cols}=1); \ident{types}(\ident{string})\\[0.2em]
\end{array}
\end{equation*}
%
At the first level, the AI assistant gives us three options. The first, most likely, one skips just 
the first line and identifies that there are three columns (perhaps because the second line contains 
two commas and can thus be interpreted as data row). The second skips first two lines and recognises
three columns. The AI assistant might find this less likely, but it is the option we want to choose
as users. Finally, the third option takes all lines, but then it has to join all columns into one
which is unlikely.

The user can choose the second option, at which point a user interface can evaluate the 
expression $\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3)$ and show a preview of 
the cleaning result. Next, the user has a choice of two ways of inferring types -- the most 
likely one (and the right one) is that the third column has a type \ident{int} (which means that
the value {\texttt na} in the last row will be treated as missing).

@@

The datadiff project takes two tabular datasets that are in inconsistent formats, but represent
two sets of observations about the same entity (such as broadband speed in two different years).
It generates a list of patches that can be applied to one dataset in order to convert it to the
format used by the second dataset. The patches form a domain-specific language (DSL) of the 
data transformation suggested by datadiff. A patch $p$ can be one of the following
(the details are not important for this document, but it provides a nice example of a DSL):

\begin{equation*}
\begin{array}{lcll}
p &=& \ident{permute}(\pi) &\textnormal{(Reorder columns using a given permutation)}\\
  &|& \ident{recode}(i, [s_1\mapsto s_2, \ldots]) &\textnormal{(Recode categorical values using in column $i$)}\\
  &|& \ident{linear}(i, a, b)&\textnormal{(Apply linear transform $va+b$ to a numerical column $i$)}\\
  &|& \ident{delete}(i)&\textnormal{(Drop the column at index $i$)}\\
  &|& \ident{insert}(i, d)&\textnormal{(Insert a new column at index $i$ initialized to values $d$)}
\end{array}
\end{equation*}
%
The current implementation of datadiff is a non-interactive R package that takes two datasets
and produces a list of patches. It takes a couple of parameters that specify which patches should
datadiff attempt to generate (together with various weights to tweak the optimization).

@@

\noindent
If we ask datadiff for a list of patches to transform the dataset on the right-hand side
to the format of the dataset on the left-hand side, we might get the following patches
(assuming columns are indexed from $1$):

\[
\ident{delete}(3);~ \ident{permute}(2,1);~ \ident{recode}(2, [\textnormal{\sffamily "\hspace{-0.2em}Cardiff\hspace{0.1em}"}\mapsto\textnormal{\sffamily "\hspace{-0.2em}London"}])
\]

@@

The datadiff AI assistant generates the following tree based on the three inferred patches 
(we omit the parameters of the patches):
%
\begin{equation*}
\begin{array}{l}
p=0.9,~ e=\ident{delete}(\ldots)\\[0.2em]
\qquad\longrightarrow p=0.9,~ \ident{delete}(\ldots); \ident{permute}(\ldots)\\[0.2em]
\qquad\qquad\longrightarrow p=0.6,~ \ident{delete}(\ldots); \ident{permute}(\ldots); \ident{recode}(\ldots)\\[0.2em]
\qquad\longrightarrow p=0.6,~ \ident{delete}(\ldots); \ident{recode}(\ldots)\\[0.2em]
\qquad\qquad\longrightarrow p=0.9,~ \ident{delete}(\ldots); \ident{recode}(\ldots); \ident{permute}(\ldots)\\[0.4em]
p=0.8,~ e=\ident{permute}(\ldots)\\[0.2em]
\qquad\longrightarrow p=0.9,~ \ident{permute}(\ldots); \ident{delete}(\ldots)\\[0.2em]
\qquad\qquad\longrightarrow p=0.6,~ \ident{permute}(\ldots); \ident{delete}(\ldots); \ident{recode}(\ldots)\\[0.2em]
\qquad\longrightarrow p=0.6,~ \ident{permute}(\ldots); \ident{recode}(\ldots)\\[0.2em]
\qquad\qquad\longrightarrow p=0.9,~ \ident{permute}(\ldots); \ident{recode}(\ldots); \ident{delete}(\ldots)\\[0.4em]
(\ldots)
\end{array}
\end{equation*}

>>>>

\documentclass{article}
\usepackage{color}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{tgtermes}
\usepackage{forest}
\usepackage{fancyvrb}
\usepackage{setspace}
\usepackage[font=small,labelfont=bf]{caption}

\forestset{
  tt/.style={
    for tree={
      parent anchor=south,
      child anchor=north,
      grow=270,
      align=center,
      inner sep=0pt
    }
  },
  nice empty nodes up/.style={
    for tree={calign=fixed edge angles},
    delay={where content={}{shape=coordinate,for parent={for children={anchor=south}}}{}}
  }
}
\begin{document}
\title{\textbf{AI assistants}: A framework for semi-automated, transparent, tooling-rich data wrangling}
\author{Tomas Petricek}
\maketitle
% \begin{abstract}
%   zz
% \end{abstract}

\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\definecolor{tpclr}{rgb}{0.6,0,0.3}
\newcommand{\tpnote}[1]{{\marginpar{
  \setstretch{0.8}\raggedright\textcolor{tpclr}{ \footnotesize{\textbf{Tomas:} #1 } }}
}}
~

\section{Introduction}
\tpnote{This part is written as an intro to a potential paper about AI assistants 
(using many ideas and formulations from Charles et al.)}

According to some reports, 80\% of data analysis is spent on \emph{data wrangling} -- an activity
that encompasses obtaining data from a diversity of semi-structured sources, making sense of data
encoding, joining multiple datasets with mismatched structure, correcting erroneous records and 
filling missing data. Despite the recent rise of artificial intelligence and machine learning, 
data wrangling has, so far, largely eluded automation and remains a tedious and time-consuming 
manual task.

There are a number of reasons why data wrangling is hard to automate. First, data is often big and 
diverse and hide hard-to-find special cases that can be only handled with some human insight.
Second, real-world data is heterogeneous. Automatic methods developed for one kind of data will 
often fail when more data is added. Third, data cleaning often requires tuning of many parameters 
and you often cannot know what will work until you review the results. Finally, making an error
during data wrangling can have a disasterous effect on a project. An automatic tool can easily 
confuse interesting outliers for an uninteresting noise in cases where a quick human glance would
immediately spot the distinction.

For these rasons, we argue that a major advance in data wrangling is best achieved by developing
semi-automatic methods that allow us to effectively combine automation and scalability of machine 
learning methods with human insight. To capture the human-computer interaction pattern that is
required for data wrangling, we introduce the concept of an \emph{AI assistant}. An AI assistant
is a component that interacts with a data analyst to guide them through a single data extraction,
collection, integration, preparation or exploration task. AI assistants are semi-automated, 
transparent and tooling-rich:
%
\begin{enumerate}
\item \textbf{Semi-automated.} After being invoked on input data, the AI assistant runs some automated
  data wrangling algorithm (based on AI, machine learning or statistical analysis) and offers a list
  of suggestions to the data analyst. The analyst can review the results, choose one of the provided
  options or, more generally, provide feedback back to the algorithm. The loop is repeated until 
  the AI assistant has no further suggestions or until the analyst accepts one of the suggested results.

\vspace{-0.2em}
\item \textbf{Transparent.} An AI assistant needs to explain what cleaning operations it recommends. 
  At each step, the result of running an AI assistant is a short data wrangling script, written 
  in a domain-specific language, designed for each individual assistant. The data analyst can 
  review the script to assess if the recommended cleaning operation is reasonable, before running
  the script on the input data.

\vspace{-0.2em}
\item \textbf{Tooling-rich.} Finally, an AI assistant needs to play well with modern data science
  tooling. This means that offering the suggestions to the data analyst should be done through an 
  interactive programming mechanism such as auto-complete. When reviewing suggestions, the data 
  analyst should also immediately see a preview of running the script on the actual input data as,
  for example, when running code in Jupyter notebooks.
\end{enumerate}
%
Several tools that satisfy the above definition of an AI assistant have been built in the past, 
but our main motivation is to describe the general structure of an AI assistant abstractly and
discuss the methodology of developing and evaluating AI assistants.

A data analyst will typically need to interact with many separate assistants over the course of an
anlysis, each of which specializes in a single kind of task. Having a clear abstract definition
of what an AI assistant is, together with several examples, will provide a common framework that 
developers of AI assistants and authors of data science tools can use in order to build a 
flourishing ecosystem for semi-automated, transparent, tooling-rich data wrangling.

\tpnote{I think this is good structure of contributions for a real paper. The current version does not quite do all this yet though!}
This paper explains what an AI assistant is, discusses how it can be integrated in a notebook
environment for data science and gives a number of examples of AI assistants: 
%
\begin{itemize}
\item We give a formal description that models AI assistants as trees (Section~\ref{sec:trees}). 
  An AI assistant offers user a choice by generating multiple sub-trees and the user makes a choice 
  and provides an input to the AI assistant by choosing one branch.
\vspace{-0.2em}
\item We discuss the integration of AI assistants in the Wrattler notebook system (Section~\ref{sec:notebook}). 
  This makes it easy to use AI assistants interactively and Wrattler also gives rapid feedback when 
  cleaning data using an AI assistant. 
\vspace{-0.2em}
\item We give two examples of AI assistants and explain how the tree-based model captures those
  examples (Section~\ref{sec:examples}). We consider \emph{Datadiff}, which automates the process of reconcilling 
  inconsistencies between pairs of tabular data sets and a \emph{CSV parser} that extracts typed
  data from real-world messy CSV files.
\vspace{-0.2em}
\item We consider two ways of evaluating AI assistants (Section~\ref{sec:evaluation}). An AI 
  assistant should provide the best answer when used fully automatically, but user interaction 
  should allow correcting it when the answer obtained automatically is not right.
\end{itemize}

\section{Background}
\label{sec:background}
\tpnote{These are comments for AIDA team readers only, somewhat overlapping with the introduction.
We'll need to rewrite this for a real paper. }

The idea of an \emph{AI assistant} appeared in many contexts in the AIDA project.
Wrattler aims to provide a platform through which AI assistants can be 
easily used; the datadiff project developed an AI assistant which automates the process of 
reconciling inconsistencies between pairs of tabular datasets. However, we never clearly defined
what an AI assistant is.

A simple way of trying to address the data wrangling problem would be to build a set of functions
that take some dirty datasets and produce one or more clean datasets. The user could then call these
functions on their dirty data and would get a clean result. The function would likely take some
parameters that specify how cleaning should be done. (Conceivably, this could be further automated
and a system could invoke such functions automatically.)

However, this simple approach has two problems. First, real-world data sets often 
contain numerous corner cases that require some human insight and so getting the right results 
might involve laborious fiddling with function parameters. Second, we want the data analysis
to be transparent and be able to understand how have the data wrangling problems been solved. 
A simple function does not address these because it does not encourage human interaction
and it is not transparent -- it produces a result without indicating how and why.

\subsection{Why are assistants more than functions}

\tpnote{I like the idea of contrasting `cleaning function' with an `assistant' but this text
repeats ideas from the intrduction and does not do a very good job overall.}
AI assistants address the above issues. First, AI assistants are interactive. Given 
some data, they perform some analysis and provide the user with several options to choose 
from (those can be most likely solutions to some problem, possible ways of resolving ambiguous
corner cases, or a choice of next steps the AI assistant can follow). In this paper, the options are 
a finite list (where the user has to choose one), but they could also be open-ended (where the user 
can specify some input). The list of options is also sorted by quality (if we always choose the 
most likely option, the assistant will behave as an automatic function).

@@

Note that each AI assistant can use its own domain-specific language -- the language should be
as small as possible, i.e. it should be able to express the data transformations that the AI
assistant can recommend, but nothing more. The combination of interactivity and domain-specific 
languages allows us to do a number of interesting things:

@@

\begin{figure}
\vspace{-1em}
\includegraphics[scale=0.29]{wrattler.png}
\caption{The datadiff AI assistant running inside Wrattler, a notebook-like environment for
data science. The completion list shows data cleaning operations offered by an AI assistant 
and the preview shows the result of running the provided cleaning script.}
\label{fig:wrattler}
\end{figure}

\section{AI assistants in a notebook environment}
\label{sec:notebook}

Before looking at the formal model which aims to explain what an AI assistant is, it is useful 
to explore how AI assistants can be used in a notebook programming environment for data science 
such as Jupyter. Figure~\ref{fig:wrattler} shows a screenshot of a datadiff AI assistant, running
in the Wrattler notebook system.

In the example, a user obtained two datasets, \texttt{bb2015} and \texttt{bb2014} that contain 
information about broadband speeds for years 2015 and 2014. The two data sets record observations
over two years, containing a common subset of data, but in a mismatched format -- the column names
and order differ. The data analyst invokes the datadiff AI 
assistant in order to turn the format of 2015 data into the format used by the 2014 dataset.
To do this, they type \texttt{datadiff.adapt(bb2015, bb2014)}, which invokes the AI assistant.
They then type ``.'' to get a data cleaning recommendation.

The AI assistant analyses the data and determines a list of patches that needs to be applied in
order to reformat the 2015 dataset. It then offers those patches in the autocompletion list, 
sorted from the most likely patch to the least likely patch. The user can proceed by selecting 
the first patch repeatedly (as in the above screenshot). After selecting a patch, the source code
in the cell changes and the preview is updated, showing the result of running the current cleaning
script. This interactivity makes it easy to see possible data cleaning errors. If applying a patch
does not lead to an expected result, the user can easily go back and select the second most likely
patch.

Modelling an AI assistant as a tree, we can see that the code entered in the editor corresponds to
a path through the tree (where the user provides information for the AI asistant; here, by 
accepting some of the recommended patches). The options shown in an auto-complete list correspond 
to the possible sub-trees in the current tree node (here, the AI assistant is asking for an user 
input or a recommendation on how to continue).

\section{Formal model of AI assistants}
\label{sec:trees}

An AI assistant is a concept that spans multiple research areas. In order to understand how to 
integrate AI assistants into programming tools such as notebooks, we need to use methods from 
programming language research. To implement interesting AI assistants, we need to use machine 
learning and statistical models.

From a programming language perspective, we are mainly interested in the abstract structure of the
AI assistant. A programming langauge model will capture how data is passed to the assistant, how
is the assistant invoked and how does it interact with the analyst. From a machine learning
perspective, we are interested in how to build AI assistants. For example, many assistants may fit
in the Bayesian learning model where the AI assistant updates an initial model -- a prior --
based on inputs from the analyst.

\tpnote{It would be nice to have more about Bayesian model of AI assistants - but I'm not qualified
to write that. We'd need to show how a Bayesian model defines ``AI assistant as a tree structure'' 
(probably quite easy) and give some Bayesian AI assistant examples... (could be quite simple, but
we don't have them)}
However, it is conceivable that an AI assistant can be implemented in other ways. It might provide
suggestions based on a simple frequentist statistical analysis, or it might use models based on 
neural networks. The programming language model of what is an AI assistant should be able to 
accommodate all those different implementations. Modelling AI assistants as trees, as discussed
in the next section, satisfies this requirement as it allows a range of different concrete
implementations.

\subsection{Programming abstraction: AI assistants as trees}
\label{sec:trees-def}

To discuss properties of AI assistants from a programming language perspective, we model an AI 
assistant as a function that takes some input data and generates a tree where each node offers a 
finite list of options that the user can choose from. For simplicity, we ignore the possibility of 
open-ended options where the user specifies some input and we also ignore the fact that the AI 
assistant might take some initial parameters in addition to the input. Assuming \ident{Data}
is a representation of input data:
%
\begin{equation*}
\begin{array}{rcll}
a &=& \ident{Data} \rightarrow t &\textnormal{(AI assistant)}\\
t &=& (n_1, r_1, e_k, t_1), \ldots (n_k, r_k, e_k, t_k) &\textnormal{(Provided options)}\\
\mathit{eval} &=& \ident{Data}\times e \rightarrow \ident{Data}&\textnormal{(Evaluator for the DSL)}
\end{array}
\end{equation*}
%
An AI assistant $a$ is a function that takes some input data and returns a root tree node~$t$. 
A tree node consists of a list of branches $(n_i, r_i, e_i, t_i)$ where $n_i$ is a name of the branch 
(shown to the user, but otherwise unimportant), $r_i$ is a number that gives a rating of the option
(between 0 and 1), $e_i$ is an expression in the domain-specific language that represents
the data transformation attached to the node. Finally, $t_i$ is a sub-tree that can be empty or can 
offer more follow-up options. Note that this sub-tree can be generated lazily -- it is only needed 
if the user chooses the option with which it is associated and so we do not need to evaluate the 
full tree upfront.

The nature of expressions $e$ depends on the particular AI assistant and so we leave that abstract
for now. However, we certainly need to be able to apply the transformation on sample dataset. 
This is captured by the $\mathit{eval}$ function, which takes data together with an expression $e$
in the domain-specific langauge and returns transformed data.

\newpage

\begin{figure}
\hspace{4em}\begin{forest}, tt  
  [
    [{$\mathbf{n_1}$\\$p=0.9$\\$e=e_1$}
      [{$\quad\mathbf{n_{11}}\quad$\\$r=0.9$\\$e=e_{11}$}]
      [{$\quad\mathbf{n_{12}}\quad$\\$r=0.7$\\$e=e_{12}$}]
    ]
    [{$\mathbf{n_2}$\\$p=0.8$\\$e=e_2$}
      [{$\quad\mathbf{n_{21}}\quad$\\$r=0.9$\\$e=e_{21}$}]
      [{$\quad\mathbf{n_{22}}\quad$\\$r=0.8$\\$e=e_{22}$}]
    ]
    [{$\mathbf{n_3}$\\$p=0.7$\\$e=e_3$}
      [{$\quad\mathbf{n_{31}}\quad$\\$r=0.8$\\$e=e_{31}$}]
      [{$\quad\mathbf{n_{32}}\quad$\\$r=0.6$\\$e=e_{32}$}]
    ]
  ]
\end{forest}
\caption{An illustration of a formal model that views AI assistants as trees.}
\label{fig:aitree}
\vspace{-1em}
\end{figure}

\noindent
Figure~\ref{fig:aitree} illustrates the formal model and shows an (abstract) example of a tree
that can be generated by an AI assistant. The root node offers three options represented by nodes 
$n_1, n_2, n_3$. The most likely one as recommended by the AI assistant is $n_1$ with rating $r=0.9$. 
The expression attached to the node is $e_1$ and when the user selects the node, we can display a 
preview using sample data $\mathit{sample}$ using the $\mathit{eval}$ function by evaluating 
$\mathit{eval}(\mathit{sample}, e_1)$.

Assuming the user chooses the most likely option $n_1$, they will then be presented with two
subsequent options, $n_{11}, n_{12}$. Even though the AI assistant marks the first node as more
likely, the user may review the previews $\mathit{eval}(\mathit{sample}, e_{11})$,
$\mathit{eval}(\mathit{sample}, e_{12})$ and decide that the latter is actually the correct
data cleaning operation. They can then choose the node $n_{12}$ as the final one use the 
cleaning script represented by the DSL expression $e_{12}$. The expression $e_{12}$ can be used
to get the clean data by evaluating $\mathit{eval}(\mathit{sample}, e_{12})$, but it can also be
translated to a programming language such as R or Python and used in a larger data wrangling 
workflow.

\section{Examples of AI assistants}
\label{sec:examples}

\tpnote{The examples here illustrate the idea, but are not detailed enough for a proper paper
(we should also have only the better version of datadiff). It would be ideal to have CSV, better
datadiff and some Bayesian example. Possibly also ptype, but it might not fit the model that well --
we need to think about that.}

We now consider two AI assistants through the lens of our abstract tree model.
The first is hypothetical AI assistant for extracting data from CSV files. The second
is the datadiff assistant. For datadiff, we discuss the current implementation in Wrattler, but
also briefly sketch a possible improvement (that improves the interactivity aspect).

@@

\tpnote{This is inspired by Gerrit's work, but it's very basic at this point.}

As an example, consider the following broken CSV file. The file starts with two meta-data lines
(one specifying the name and another specifying copyright information). This is followed by headers
and actual data.

@@

A minimalistic AI assistant for cleaning CSV files generates tree with two levels of nodes. 
In the first level, it infers where the actual data is in the CSV file (and it offers 
its best guesses to the user). In the second level, it infers the types of the columns.

The domain-specific language for CSV parsing consists of just two functions named
$\ident{range}$ and $\ident{types}$ that specify the range of data in the file and the types
of the columns: $\ident{range}(\ident{lines}=l_n\ldots l_m, \ident{cols}=k)$ denotes that
the data starts on line $l_n$ and ends on line $l_m$ and there are $k$ columns; 
$\ident{types}(t_1,\ldots, t_k)$ specifies that the types of the $k$ columns are 
$t_1, \ldots, t_k$. The types are either $\ident{string}$ or $\ident{int}$.

\begin{figure}
\begin{equation*}
\begin{array}{l}
r=0.9,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3)\\[0.2em]
\qquad\longrightarrow r=0.9,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{string})\\[0.2em]
\qquad\longrightarrow r=0.3,~ e=\ident{range}(\ident{lines}=2\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{int})\\[0.4em]
r=0.7,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3)\\[0.2em]
\qquad\longrightarrow r=0.8,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{int})\\[0.2em]
\qquad\longrightarrow r=0.7,~ e=\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3); \ident{types}(\ident{string},\ident{string},\ident{string})\\[0.4em]
r=0.3,~ e=\ident{range}(\ident{lines}=1\ldots 6, \ident{cols}=1)\\[0.2em]
\qquad\longrightarrow r=0.9,~ e=\ident{range}(\ident{lines}=1\ldots 6, \ident{cols}=1); \ident{types}(\ident{string})\\[0.2em]
\end{array}
\end{equation*}
\caption{A tree generated by an AI assistant for parsing CSV files. Each node has a rating $r$ and
a data cleaning expression $e$ written in a small domain-specific language for cleaning CSV files.}
\label{fig:csvtree}
\end{figure}

Given the above CSV file as an input, the AI assistant produces a tree shown in Figure~\ref{fig:csvtree}.
At the first level, the AI assistant gives us three options. The first, most likely, one skips just 
the first line and identifies that there are three columns (this error is made because the second line contains 
two commas and can thus be interpreted as data row). The second option skips first two lines and recognises
three columns. The AI assistant might find this less likely, but it is the option we want to choose
as users. Finally, the third option takes all lines, but then it has to join all columns into one
which is unlikely.

The user can choose the second option, at which point a user interface can evaluate the 
expression $\ident{range}(\ident{lines}=3\ldots 6, \ident{cols}=3)$ and show a preview of 
the cleaning result as a parsed table, but without types. Next, the user has a choice of two ways of inferring types -- the most 
likely one (and the right one) is that the third column has a type \ident{int} (which means that
the value {\texttt na} in the last row will be treated as missing). If {\texttt na} was not 
representing a missing value, the user could choose the second option which infers the type of the 
last column as \ident{string}.

@@

\tpnote{I describe current version of datadiff first and then a possible improvement. If we turn
this into a paper, we should have just one (preferably the latter) version. This section might
be useful for the AIDA team though.}
The datadiff package takes two tabular datasets that are in inconsistent formats, but represent
two sets of observations about the same entity. An example is measurements of broadband speed in
two different yers, stored as two tabular files with missing or additional columns, using different
order of columns and different column names.

Datadiff generates a list of patches that can be applied to one dataset in order to convert it to the
format used by the second dataset. The patches form a domain-specific language (DSL) of the 
data transformation suggested by datadiff. A patch $p$ can be one of the following:
%
\begin{equation*}
\begin{array}{lcll}
p &=& \ident{permute}(\pi) &\textnormal{(Reorder columns using a given permutation)}\\
  &|& \ident{recode}(i, [s_1\mapsto s_2, \ldots]) &\textnormal{(Recode categorical values using in column $i$)}\\
  &|& \ident{linear}(i, a, b)&\textnormal{(Apply linear transform $va+b$ to a numerical column $i$)}\\
  &|& \ident{delete}(i)&\textnormal{(Drop the column at index $i$)}\\
  &|& \ident{insert}(i, d)&\textnormal{(Insert a new column at index $i$ initialized to values $d$)}
\end{array}
\end{equation*}
%
The current implementation of datadiff is a non-interactive R package that takes two datasets
and produces a list of patches. It takes a couple of parameters that specify which patches should
datadiff attempt to generate (together with various weights to tweak the optimization).

@@

\noindent
If we ask datadiff for a list of patches to transform the dataset on the right-hand side
to the format of the dataset on the left-hand side, we might get the following patches
(assuming columns are indexed from $1$):
%
\[
\ident{delete}(3);~ \ident{permute}(2,1);~ \ident{recode}(2, [\textnormal{\sffamily "\hspace{-0.2em}Cardiff\hspace{0.1em}"}\mapsto\textnormal{\sffamily "\hspace{-0.2em}London"}])
\]

@@

\begin{figure}
\begin{equation*}
\begin{array}{l}
r=0.9,~ e=\ident{delete}(\ldots)\\[0.2em]
\qquad\longrightarrow r=0.9,~ \ident{delete}(\ldots); \ident{permute}(\ldots)\\[0.2em]
\qquad\qquad\longrightarrow r=0.6,~ \ident{delete}(\ldots); \ident{permute}(\ldots); \ident{recode}(\ldots)\\[0.2em]
\qquad\longrightarrow r=0.6,~ \ident{delete}(\ldots); \ident{recode}(\ldots)\\[0.2em]
\qquad\qquad\longrightarrow r=0.9,~ \ident{delete}(\ldots); \ident{recode}(\ldots); \ident{permute}(\ldots)\\[0.4em]
r=0.8,~ e=\ident{permute}(\ldots)\\[0.2em]
\qquad\longrightarrow r=0.9,~ \ident{permute}(\ldots); \ident{delete}(\ldots)\\[0.2em]
\qquad\qquad\longrightarrow r=0.6,~ \ident{permute}(\ldots); \ident{delete}(\ldots); \ident{recode}(\ldots)\\[0.2em]
\qquad\longrightarrow r=0.6,~ \ident{permute}(\ldots); \ident{recode}(\ldots)\\[0.2em]
\qquad\qquad\longrightarrow r=0.9,~ \ident{permute}(\ldots); \ident{recode}(\ldots); \ident{delete}(\ldots)\\[0.4em]
(\ldots)
\end{array}
\end{equation*}
\caption{A tree generated by a basic datadiff AI assistant. The parameters of the patches are
omitted -- they are the same as in the full list of patches shown above. }
\label{fig:ddtree}
\end{figure}
